Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-27 article-models-tt.tex
judgmental pullbacks

\subsection{Extension on terms}\label{extension-on-terms}
In this subsection we consider the notion of extension on terms, which has now
become definable inside our theory. Moreover, every compatibility rule one may
dream of is provable as well, using the compatibility rules we have introduced
earlier. \emph{This subsection contains no new assumptions.}

@@

\subsubsection{Commuting dependence diagrams}
Now that we have introduced the notions of morphisms, composition and identity
terms, we can develop a diagramatic style of of displaying type dependencies
combined with morphisms. We give an informal, metatheoretical definition of
such diagrams by indicating what the various components mean. The definition
is informal because we will only use such diagrams occasionally to provide a
graphical indication of the situation in which we're working.

\begin{defn}
A diagram is said to be a \emph{dependency diagram in context $\Gamma$}
if it is built up according to the following steps:
\begin{itemize}
\item An ordinary arrow 
\begin{equation*}
\begin{tikzcd}
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
between two families $A$ and $B$ of contexts over $\Gamma$ indicates that
$f$ is a morphism from $A$ to $B$ in context $\Gamma$, i.e.~that we have the
judgment $\jhom{\Gamma}{A}{B}{f}$.
\item The set of double-headed arrows must form a forest where the root of
each tree is a family of contexts over $\Gamma$. 
\item A sequence of double-headed 
arrows
\begin{equation*}
\begin{tikzcd}
P_{n+1} \ar[fib]{r} & \cdots \ar[fib]{r} & P_1 \ar[fib]{r} & A
\end{tikzcd}
\end{equation*}
indicates that $P_1$ is a family of contexts over $\ctxext{\Gamma}{A}$, that
$P_2$ is a family of contexts over $\ctxext{{\Gamma}{A}}{P_1}$, etcetera.
\item A ladder of double-headed arrows
\begin{equation*}
\begin{tikzcd}
P_{n+1} \ar{r}{F_{n+1}} \ar[fib]{d} & Q_{n+1} \ar[fib]{d}\\
\vdots \ar[fib]{d} & \vdots \ar[fib]{d}\\
P_1 \ar{r}{F_1} \ar[fib]{d} & Q_1 \ar[fib]{d}\\
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
indicates that $F_1$ is a morphism from $P_1$ to $Q_1$ \emph{over} $f$,
i.e.~that the judgment $\jfhom{\Gamma}{A}{B}{f}{P_1}{Q_1}{F_1}$ holds, that
$F_2$ is a morphism from $P_2$ to $Q_2$ over
the morphism $\tmext{}{}{\ctxwk{P_1}{f}}{F_1}$ from $\ctxext{A}{P_1}$ to
$\ctxext{B}{Q_1}$, etcetera.
\end{itemize}
Such a diagram is said to be commutative if the subdiagram consisting of only
the normal headed arrows is commutative in the usual sense (using judgmental
equality). Note that the ladders are inherently commutative.
\end{defn}

@@

\subsection{Judgmental trivial cofibrations}
\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a context morphism. We say that $f$ is a 
\emph{judgmental trivial cofibration} if the following inference rules are valid:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jterm{{\Gamma}{B}}{Q}{\jtcext{g}}
    }
& & \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jtermeq{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{\jcomp{A}{f}{\jtcext{g}}}{g}
    }
\end{align*}
\end{defn}

>>>>

\subsection{Digging a little deeper in the theory of contexts, families and terms}

There is a good amount of things we can already state and prove about in the
theory we have developed so far. We will develop a bit more theory that will
come in handy later on.

\emph{This subsection contains no new assumptions.}

\subsubsection{Extension on terms}\label{extension-on-terms}
In this subsubsection we consider the notion of extension on terms, which has now
become definable inside our theory. Moreover, every compatibility rule one may
dream of is provable as well, using the compatibility rules we have introduced
earlier.

@@

\subsubsection{Commuting dependence diagrams}
Now that we have introduced the notions of morphisms, composition and identity
terms, we can develop a diagramatic style of of displaying type dependencies
combined with morphisms. We give an informal, metatheoretical definition of
such diagrams by indicating what the various components mean. The definition
is informal because we will only use such diagrams occasionally to provide a
graphical indication of the situation in which we're working. In particular,
we will not shy away from using natural numbers and trust that the reader can
figure out what we mean.

\begin{defn}
A diagram is said to be a \emph{dependency diagram in context $\Gamma$}
if it is built up according to the following steps:
\begin{itemize}
\item The arrows appearing in a dependency diagram are either ordinary, like the
arrow%
$\begin{tikzcd}[ampersand replacement = \&]
X \ar{r} \& Y,
\end{tikzcd}$
or double-headed, like
$\begin{tikzcd}[ampersand replacement = \&]
X \ar[fib]{r} \& Y.
\end{tikzcd}$
\item An ordinary arrow 
\begin{equation*}
\begin{tikzcd}
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
between two families $A$ and $B$ of contexts over $\Gamma$ indicates that
$f$ is a morphism from $A$ to $B$ in context $\Gamma$, i.e.~that we have the
judgment $\jhom{\Gamma}{A}{B}{f}$.
\item The set of double-headed arrows must form a forest and the root of
each maximal tree of double-headed arrows is a family of contexts over $\Gamma$.
In particular, if an object is not the domain of a double-headed arrow it must
be a family of contexts over $\Gamma$.
\item A sequence of double-headed 
arrows
\begin{equation*}
\begin{tikzcd}
P_{n} \ar[fib]{r} & \cdots \ar[fib]{r} & P_1 \ar[fib]{r} & A
\end{tikzcd}
\end{equation*}
indicates that $P_1$ is a family of contexts over $\ctxext{\Gamma}{A}$, that
$P_2$ is a family of contexts over $\ctxext{{\Gamma}{A}}{P_1}$, etcetera.
\item There can be two kinds of ladders of double-headed arrows:
\begin{equation*}
\begin{tikzcd}
P_{n} \ar{r}{F_{n}} \ar[fib]{d} & Q_{n} \ar[fib]{d}\\
\vdots \ar[fib]{d} & \vdots \ar[fib]{d}\\
P_1 \ar{r}{F_1} \ar[fib]{d} & Q_1 \ar[fib]{d}\\
A \ar{r}{f} & B
\end{tikzcd}
\qquad
\begin{tikzcd}[column sep = tiny]
P_{n+m} \ar{rr}{F_{n+m}} \ar[fib]{d} & & Q_{n+m} \ar[fib]{d}\\
\vdots \ar[fib]{d} & & \vdots \ar[fib]{d}\\
P_{n+1} \ar{rr}{F_{n+1}} \ar[fib]{dr} & & Q_{n+1} \ar[fib]{dl}\\
& P_n \ar[fib]{d}\\
& \vdots \ar[fib]{d}\\
& P_1 \ar[fib]{d}\\
& A
\end{tikzcd}
\end{equation*}
The ladder on the left 
indicates that $F_1$ is a morphism from $P_1$ to $Q_1$ \emph{over} $f$,
i.e.~that the judgment $\jfhom{\Gamma}{A}{B}{f}{P_1}{Q_1}{F_1}$ holds, that
$F_2$ is a morphism from $P_2$ to $Q_2$ over
the morphism $\tmext{}{}{\ctxwk{P_1}{f}}{F_1}$ from $\ctxext{A}{P_1}$ to
$\ctxext{B}{Q_1}$, etcetera.

The ladder on the right indicates that $F_{n+1}$ is a morphism from $P_{n+1}$ to
$Q_{n+1}$ in the appropriate context, that $F_{n+2}$ is a morphism from
$P_{n+2}$ to $Q_{n+2}$ over $F_{n+1}$, etcetera.

Note that the object(s) at the bottom of a ladder are always families of contexts
over $\Gamma$, so that the typing of the various ingredients makes sense.
\end{itemize}
Such a diagram is said to be commutative if the subdiagram consisting of only
the normal headed arrows is commutative in the usual sense (using judgmental
equality). Note that the ladders are inherently commutative.
\end{defn}

The most basic illustrative example of a commutative dependency diagram is
the diagram
\begin{equation*}
\begin{tikzcd}
P \ar[fib]{d} \ar{r}{F} & Q \ar[fib]{d} \\
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
indicating a morphism $F$ from $P$ to $Q$ over the morphism $f$ from $A$ to
$B$ in a context $\Gamma$.

\begin{defn}
We say that a commutative dependency diagram of the form
\begin{equation*}
\begin{tikzcd}
P \ar[fib]{d} \ar{r}{F} & Q \ar[fib]{d} \\
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
is a judgmental pullback if the following inference rules are valid:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{A}}{P'}
    \jfhom{\Gamma}{A}{B}{f}{P'}{Q}{F'}
    }
  { \jhom{{\Gamma}{A}}{P'}{P}{u}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P'}
    \jfhom{\Gamma}{A}{B}{f}{P'}{Q}{F'}
    }
  { \jfhomeq{\Gamma}{A}{B}{f}{P'}{Q}{\jcomp{}{u}{F}}{F'}
    }
  \\
& \inference
  { \jhom{{\Gamma}{A}}{P'}{P}{v}
    \jfhomeq{\Gamma}{A}{B}{f}{P'}{Q}{\jcomp{}{v}{F}}{F'}
    }
  { \jhomeq{{\Gamma}{A}}{P'}{P}{v}{u}
    }
\end{align*}
\end{defn}

\begin{lem}
The diagram
\begin{equation*}
\begin{tikzcd}
\jcomp{}{f}{Q} \ar[fib]{d} \ar{r}{\idtm{\jcomp{}{f}{Q}}} & Q \ar[fib]{d} \\
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
is a judgmental pullback diagram.
\end{lem}

@@

\subsubsection{Judgmental trivial cofibrations}
\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a context morphism. We say that $f$ is a 
\emph{judgmental trivial cofibration} if the following inference rules are valid:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jterm{{\Gamma}{B}}{Q}{\jtcext{g}}
    }
& & \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jtermeq{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{\jcomp{A}{f}{\jtcext{g}}}{g}
    }
\end{align*}
\end{defn}

### 2014-03-27 article-models-ttconstr.tex
judgmental pullbacks

\subsection{Universes}

>>>>

\subsection{Universes}
The type operator $\UU$ has as input a pair $(A,P)$ consisting of a family
$A$ in context $\Gamma$ and a family $P$ in context $\ctxext{\Gamma}{A}$. Its
output is a pair $(U(A,P),\tilde{U}(A,P))$ consisting of a family
$U(A,P)$ in context $\Gamma$ and a family $\tilde{U}(A,P)$ over it for which
there are terms representing the empty context, the operations of extension,
weakening and substitution and the identity terms. Moreover there is a morphism
$i$ from $A$ to $U(A,P)$ with the property that 
$\jcomp{}{i}{\tilde{U}(A,P)}\jdeq P$.
